---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 4,
  fig.align = "center"
)
```

[\@drsimonj](https://twitter.com/drsimonj) here to discuss the latest CRAN release of **corrr** (0.2.1), a package for exploring **corr**elations in a tidy **R** framework. This post will describe corrr features added since version 0.1.0.

You can install or update to this latest version directly from CRAN by running:

```{r eval = FALSE}
install.packages(corrr)
```

Let's load corrr into our workspace and create a correlation data frame of the mtcars data set to work with:

```{r, message = F}
library(corrr)
rdf <- correlate(mtcars)
rdf
```

## Plotting functions

The major changes involve the `rplot()` and new `network_plot()` functions that support the visualisation of your correlations.

### rplot()

`rplot()` produces a shape plot of the correlations. More visible dots correspond to stronger correlations, and blue and red respectively to positive and negative. The default plot looks like this:

```{r rplot1}
rplot(rdf)
```

There are now four arguments that allow you to make adjustments to this plot:

- `legend` Boolean indicating whether a legend mapping the colours to the correlations should be displayed.
- `shape` geom_point aesthetic. A number correspoding to the shape to plot for each point. See <http://sape.inf.usi.ch/quick-reference/ggplot2/shape>
- `colours` or `colors` Vector of colours to use for n-colour gradient. See <http://sape.inf.usi.ch/quick-reference/ggplot2/colour>
- `print_cor` Boolean indicating whether the correlations should be printed over the shapes.

Here are some examples that change these values:

```{r rplot-shape, fig.width = 5}
rplot(rdf, legend = TRUE, shape = 1)
```

```{r rplot-color, fig.width = 5}
rplot(rdf, legend = TRUE, colours = c("firebrick1", "black", "darkcyan"))
```

```{r rplot-print}
rplot(rdf, print_cor = TRUE)
```

And don't forget that you can `rearrange()` your correlations first:

```{r rplot-rear, fig.width = 5}
rdf %>% rearrange(absolute = FALSE) %>% rplot(shape = 0, print_cor = TRUE)
```

### network_plot()

`network_plot()` produces a network that lays out and connects variables based on the strength of their correlations:

```{r netplot1}
network_plot(rdf)
```

For a good intro to `network_plot()`, see my [previous blogR post](https://drsimonj.svbtle.com/correlation-network-plot-with-corrr). There are three arguments that allow you to adjust this plot:

- `min_cor` Number from 0 to 1 indicating the minimum value of correlations (in absolute terms) to plot.
- `legend` same as `rplot()`
- `colours` or `colors` same as `rplot()`

Some examples:

```{r netplot-egs}
network_plot(rdf, legend = TRUE, colours = c("slategrey", "palegreen"))

network_plot(rdf, legend = TRUE, min_cor = .7)
```

## Other features

### fashion()

`fashion()` will now try to work on almost any object (not just correlation data frames). It also provides custom arguements to adjust the number of `decimals`, whether to display `leading_zeros`, and how to print missing values (`na_print`):

```{r}
fashion(rdf)

fashion(mtcars) %>% head()

fashion(c(0.340823, NA, -10.000032), decimals = 3, na_print = "MISSING")

fashion(c(0.340823, NA, -10.000032), leading_zeros = TRUE)
```

### focus()

A standard evaluation version of `focus()` is now available, `focus_()`, to programatically focus on specific correlations:

```{r}
vars <- c("mpg", "disp")
focus_(rdf, "hp", .dots = vars)
```

### Bugs and stuff

Other than these, there have been fixes to various bugs and minor improvements made to existing functions. Please don't forget to [open an issue on GitHub](https://github.com/drsimonj/corrr/issues) or email me if you spot an issue or would like a new feature when using corrr.

## Acknowledgements

Many thanks to the community who have already been using corrr and made suggestions along the way. Your help is invaluable for improving corrr!

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).