---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

Tired of trying to get your data to print right or formatting it in a program like excel? Try out `fashion()` from the `corrr` package:


```{r init-example, message = FALSE}
d <- data.frame(
  gender = factor(c("Male", "Female", NA)),
  age    = c(NA, 28.1111111, 74.3),
  height = c(188, NA, 168.78906),
  fte    = c(NA, .78273, .9)
)
d

library(corrr)
fashion(d)
```

But how does it work and what does it do?

## The inspiration: correlations and decimals

The insipration for `fashion()` came from my unending frustration at getting a correlation matrix to print out exactly how I wanted. For example, printing correlations typically looks something like:

```{r}
mtcars %>% correlate()
```

But this is just plain ugly. Personally, I wanted:

- Decimal places rounded to the same length (usually 2)
- All the leading zeros removed, but keeping the decimal aligned with/without `-` for negative numbers.
- Missing values (`NA`) to appear empty (`""`).

This is exactly what fashion does:

```{r}
mtcars %>% correlate() %>% fashion()
```

And if I want to change the number of `decimal` places and have a different place holder for `NA` values (`na_print`):

```{r}
mtcars %>% correlate() %>% fashion(decimals = 1, na_print = "x")
```

## Look but don't touch

There's a little bit of magic going on here, but the point to know is that `fashion()` is returning a noquote version of the original structure:

```{r}
mtcars %>% correlate() %>% fashion() %>% class()
```

That means that numbers are no longer numbers.

```{r}
mtcars %>% correlate() %>% sapply(is.numeric)

mtcars %>% correlate() %>% fashion() %>% sapply(is.numeric)
```

Similarly, missing values are no longer missing values.

```{r}
mtcars %>% correlate() %>% sapply(function(i) sum(is.na(i)))

mtcars %>% correlate() %>% fashion() %>% sapply(function(i) sum(is.na(i)))
```

So `fashion()` is for looking at output, not for continuing to work with it.

## What to use it on

`fashion()` can be used on most standard R structures such as scalars, vectors, matrices, data frames, etc:

```{r}
fashion(10.277)
fashion(c(10.3785, NA, 87))
fashion(matrix(1:4, nrow = 2))
```

You can also use it on non-numeric data. In this case, all `fashion()` will do is convert the data to characters, and then alter missing values:

```{r}
fashion("Hello")
fashion(c("Hello", NA), na_print = "World")
```

Now is a good time to take a look back at the opening example to see that it works on a data frame and with a factor column.

## Exporting

Don't forget that it's easy to export your fashioned output with something like:

```{r, eval = FALSE}
my_data %>% fashion() %>% write.csv("fashioned_file.csv")
```

So what are you waiting for? Go forth and `fashion()`!

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).
