---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 4,
  fig.align = "center",
  fig.ext = "png"
)
```

[\@drsimonj](https://twitter.com/drsimonj) here with five simple tricks I find myself sharing all the time with fellow R users to improve their code!

## 1. More fun to sequence from 1

Next time you use the colon operator to create a sequence from 1 like `1:n`, try `seq()`.

```{r eg-seq, message = FALSE}
# Sequence a vector
x <- runif(10)
seq(x)

# Sequence an integer
seq(nrow(mtcars))
```

The colon operator can produce unexpected results that can create all sorts of problems without you noticing! Take a look at what happens when you want to sequence the length of an empty vector:

```{r}
# Empty vector
x <- c()

1:length(x)

seq(x)
```

You'll also notice that this saves you from using functions like `length()`. When applied to an object of a certain length, `seq()` will automatically create a sequence from 1 to the length of the object.

## 2. `vector()` what you `c()`

Next time you create an empty vector with `c()`, try to replace it with `vector("type", length)`.

```{r}
# A numeric vector with 5 elements
vector("numeric", 5)

# A character vector with 3 elements
vector("character", 3)
```

Doing this improves memory usage and increases speed! You often know upfront what type of values will go into a vector, and how long the vector will be. Using `c()` means R has to **slowly** work both of these things out. So help give it a boost with `vector()`!

A good example of this value is in a for loop. People often write loops by declaring an empty vector and growing it with `c()` like this:

```{r, eval = FALSE}
x <- c()
for (i in seq(5)) {
  x <- c(x, i)
}
```
```{r, echo = FALSE}
x <- c()
for (i in seq(5)) {
  x <- c(x, i)
  message("x at step ", i," : ", paste(x, collapse = ", "))
}
```

Instead, pre-define the type and length with `vector()`, and reference positions by index, like this:

```{r, eval = FALSE}
n <- 5
x <- vector("integer", n)
for (i in seq(n)) {
  x[i] <- i
}
```
```{r, echo = FALSE}
n <- 5
x <- vector("integer", n)
for (i in seq(n)) {
  x[i] <- i
  message("x at step ", i," : ", paste(x, collapse = ", "))
}
```

Here's a quick speed comparison:

```{r}
n <- 1e5

x_empty <- c()
system.time(for(i in seq(n)) x_empty <- c(x_empty, i))

x_zeros <- vector("integer", n)
system.time(for(i in seq(n)) x_zeros[i] <- i)
```

That should be convincing enough!

## 3. Ditch the `which()`

Next time you use `which()`, try to ditch it! People often use `which()` to get indices from some boolean condition, and then select values at those indices. This is not necessary.

Getting vector elements greater than 5:

```{r}
x <- 3:7

# Using which (not necessary)
x[which(x > 5)]

# No which
x[x > 5]
```

Or counting number of values greater than 5:

```{r}
# Using which
length(which(x > 5))

# Without which
sum(x > 5)
```

Why should you ditch `which()`? It's often unnecessary and boolean vectors are all you need.

For example, R lets you select elements flagged as `TRUE` in a boolean vector:

```{r}
condition <- x > 5
condition
x[condition]
```

Also, when combined with `sum()` or `mean()`, boolean vectors can be used to get the count or proportion of values meeting a condition:

```{r}
sum(condition)
mean(condition)
```

`which()` tells you the indices of TRUE values:

```{r}
which(condition)
```

And while the results are not wrong, it's just not necessary. For example, I often see people combining `which()` and `length()` to test whether any or all values are TRUE. Instead, you just need `any()` or `all()`:

```{r}
x <- c(1, 2, 12)

# Using `which()` and `length()` to test if any values are greater than 10
if (length(which(x > 10)) > 0)
  print("At least one value is greater than 10")

# Wrapping a boolean vector with `any()`
if (any(x > 10))
  print("At least one value is greater than 10")

# Using `which()` and `length()` to test if all values are positive
if (length(which(x > 0)) == length(x))
  print("All values are positive")

# Wrapping a boolean vector with `all()`
if (all(x > 0))
  print("All values are positive")
```

Oh, and it saves you a little time...

```{r}
x <- runif(1e8)

system.time(x[which(x > .5)])

system.time(x[x > .5])
```

## 4. `factor` that factor!

Ever removed values from a factor and found you're stuck with old levels that don't exist anymore? I see all sorts of creative ways to deal with this. The simplest solution is often just to wrap it in `factor()` again.

This example creates a factor with four levels (`"a"`, `"b"`, `"c"` and `"d"`):

```{r factor-1}
# A factor with four levels
x <- factor(c("a", "b", "c", "d"))
x

plot(x)
```

If you drop all cases of one level (`"d"`), the level is still recorded in the factor:

```{r factor-2}
# Drop all values for one level
x <- x[x != "d"]

# But we still have this level!
x

plot(x)
```

A super simple method for removing it is to use `factor()` again:

```{r factor-3}
x <- factor(x)
x

plot(x)
```

This is typically a good solution to a problem that gets a lot of people mad. So save yourself a headache and `factor` that factor!

## 5. First you get the `$`, then you get the power

Next time you want to extract values from a `data.frame` column where the rows meet a condition, specify the column with `$` before the rows with `[`.

#### Examples

Say you want the horsepower (`hp`) for cars with 4 cylinders (`cyl`), using the `mtcars` data set. You can write either of these:

```{r}
# rows first, column second - not ideal
mtcars[mtcars$cyl == 4, ]$hp

# column first, rows second - much better
mtcars$hp[mtcars$cyl == 4]
```

The tip here is to use the second approach.

But why is that?

First reason: do away with that pesky comma! When you specify rows before the column, you need to remember the comma: `mtcars[mtcars$cyl == 4`**,**` ]$hp`. When you specify column first, this means that you're now referring to a vector, and don't need the comma!

Second reason: speed! Let's test it out on a larger data frame:

```{r}
# Simulate a data frame...
n <- 1e7
d <- data.frame(
  a = seq(n),
  b = runif(n)
)

# rows first, column second - not ideal
system.time(d[d$b > .5, ]$a)

# column first, rows second - much better
system.time(d$a[d$b > .5])
```

Worth it, right?

Still, if you want to hone your skills as an R data frame ninja, I suggest learning `dplyr`. You can get a good overview on the [`dplyr` website](http://dplyr.tidyverse.org/) or really learn the ropes with online courses like DataCamp's [Data Manipulation in R with `dplyr`](https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial).

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).
