---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 7
)
```

SOLVED: get the correlations of one or more variables with many others.

```{r init-example, message = FALSE}
library(corrr)
mtcars %>% correlate() %>% focus(mpg)
```

Let's break it down.

## Motivation

I've noticed a lot of people asking how to do this: see [here](http://stackoverflow.com/questions/38660129/how-to-correlate-and-visualise-correlation-of-one-variable-versus-many/38669144#38669144), [here](http://stackoverflow.com/questions/38548943/correlation-between-multiple-variables-of-a-data-frame/38558925#38558925), [here](http://stackoverflow.com/questions/38331537/r-using-cor-test-on-multiple-columns-on-a-data-frame/38559334#38559334). 

So this post will explain how to use `focus()` from the corrr package to correlate one or more variables in a data frame with many others.

## Starting with corrr

We'll be using the [corrr](https://cran.rstudio.com/web/packages/corrr/index.html) package, which starts by using `correlate()` to create a correlation data frame. For example, we can `correlate()` all columns in the `mtcars` data set as follows:

```{r}
mtcars %>% correlate()
```

## Introducing `focus()`

Once we have a correlation data frame, we can `focus()` on subsections of these results. For example, to `focus()` on the correlations of the `mpg` variable with all other variables, we do:

```{r}
mtcars %>% correlate() %>% focus(mpg)
```

## How does it work?

`focus()` works similarly to `select()` from the [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) package (which is loaded along with the corrr package). You add the names of the columns you wish to keep in your correlation data frame. Extending `select()`, `focus()` will then remove the remaining column variables from the rows. This is why `mpg` does not appear in the rows above. Here's another example with two variables:

```{r}
mtcars %>% correlate() %>% focus(mpg, disp)
```

`focus()` gives us all the same syntax as `select()`. For example, to drop a column (instead of keeping it), we use `-`, which will then keep it in the rows:

```{r}
mtcars %>% correlate() %>% focus(-mpg)
```

Or use utility functions like `contains()`:

```{r}
iris[-5] %>% correlate() %>% focus(contains("Sepal"))
```

For the full list of uses, explore `?select`.

## Working with the results

Let's take the case of correlating one variable with all others (e.g., `mpg` with all others). What should we do? Well, we have a data_frame with two columns:

- `rowname` which has the names of the other variables.
- `mpg` which has the correlations of `mpg` with the other variables.

With this in mind, one thing we can do is describe various aspects of the correlations within `summarise()` (also from dplyr package):

```{r}
mtcars %>% correlate() %>% focus(mpg) %>% summarise(
  n    = n(),
  mean = mean(mpg),
  sd   = sd(mpg),
  min  = min(mpg),
  max  = max(mpg)
)
```

I also like to visualise the correlations. For example, using the [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html) package:

```{r plot1, warning = F}
library(ggplot2)
mtcars %>% correlate() %>% focus(mpg) %>%
  ggplot(aes(x = rowname, y = mpg)) +
    geom_bar(stat = "identity") +
    ylab("Correlation with mpg") +
    xlab("Variable")
```

To add to this, we can order `rowname` by the correlation size in `mutate()` below:

```{r plot2, warning = F}
mtcars %>% correlate() %>% focus(mpg) %>%
  mutate(rowname = factor(rowname, levels = rowname[order(mpg)])) %>%
  ggplot(aes(x = rowname, y = mpg)) +
    geom_bar(stat = "identity") +
    ylab("Correlation with mpg") +
    xlab("Variable")
```

## Extra features

There are a few additional features to `focus()` that might help you on occasions.

First, you can keep the remaining columns in the rows (and remove all others) with the `mirror = TRUE` argument:

```{r}
mtcars %>% correlate() %>% focus(mpg:drat, mirror = TRUE)
```

For programmers, there's also a standard evaluation version `focus_()`, which behaves like `select_()`.

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).