---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  #fig.height = 3,
  #fig.width = 4,
  fig.align = "center",
  fig.ext = "png",
  dpi = 300
)
```

[\@drsimonj](https://twitter.com/drsimonj) here with a quick share on making great use of the secondary y axis with ggplot2 -- super helpful if you're plotting groups of time series!

Here's an example of what I want to show you how to create (pay attention to the numbers of the right):

```{r init-example, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)

d <- Orange %>% 
  as_tibble()

d_ends <- d %>% 
  group_by(Tree) %>% 
  top_n(1, age) %>% 
  pull(circumference)

d %>% 
  ggplot(aes(age, circumference, color = Tree)) +
    geom_line(size = 2, alpha = .8) +
    theme_minimal() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(sec.axis = sec_axis(~ ., breaks = d_ends)) +
    ggtitle("Orange trees getting bigger with age",
            subtitle = "Based on the Orange data set in R") +
    labs(x = "Days old", y = "Circumference (mm)", caption = "Plot by @drsimonj")
```

## Setup

To setup we'll need the tidyverse package and the `Orange` data set that comes with R. This tracks the circumference growth of five orange trees over time.

```{r}
library(tidyverse)

d <- Orange

head(d)
```

## Template code

To create the basic case where the numbers appear at the end of your time series lines, your code might look something like this:

```{r, eval=FALSE}
# You have a data set with:
# - GROUP colum
# - X colum (say time)
# - Y column (the values of interest)
DATA_SET

# Create a vector of the last (furthest right) y-axis values for each group
DATA_SET_ENDS <- DATA_SET %>% 
  group_by(GROUP) %>% 
  top_n(1, X) %>% 
  pull(Y)

# Create plot with `sec.axis`
ggplot(DATA_SET, aes(X, Y, color = GROUP)) +
    geom_line() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(sec.axis = sec_axis(~ ., breaks = DATA_SET_ENDS))
```

## Let's see it!

Let's break it down a bit. We already have our data set where the group colum is `Tree`, the X value is `circumference`, and the Y value is `age`.

So first get a vector of the last (furthest right) values for each group:

```{r}
d_ends <- d %>% 
  group_by(Tree) %>% 
  top_n(1, age) %>% 
  pull(circumference)

d_ends
```

Next, let's set up the basic plot without the numbers to see how each layer adds up.

```{r}
ggplot(d, aes(age, circumference, color = Tree)) +
      geom_line()
```

Now we can use `scale_y_*`, with the argument `sec.axis` to create a second axis on the right, with numbers to be displayed at `breaks`, defined by our vector of line ends:

```{r}
ggplot(d, aes(age, circumference, color = Tree)) +
      geom_line() +
      scale_y_continuous(sec.axis = sec_axis(~ ., breaks = d_ends))
```

This is a great start, The only major addition I suggest is expanding the margins of the x-axis so the gap disappears. You do this with `scale_x_*` and the `expand` argument:

```{r}
ggplot(d, aes(age, circumference, color = Tree)) +
      geom_line() +
      scale_y_continuous(sec.axis = sec_axis(~ ., breaks = d_ends)) +
      scale_x_continuous(expand = c(0, 0))
```

## Polishing it up

Like it? Here's the code to recreate the first polished plot:

```{r}
library(tidyverse)

d <- Orange %>% 
  as_tibble()

d_ends <- d %>% 
  group_by(Tree) %>% 
  top_n(1, age) %>% 
  pull(circumference)

d %>% 
  ggplot(aes(age, circumference, color = Tree)) +
    geom_line(size = 2, alpha = .8) +
    theme_minimal() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(sec.axis = sec_axis(~ ., breaks = d_ends)) +
    ggtitle("Orange trees getting bigger with age",
            subtitle = "Based on the Orange data set in R") +
    labs(x = "Days old", y = "Circumference (mm)", caption = "Plot by @drsimonj")
```


## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).