---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 4,
  fig.width = 6,
  fig.align = "center",
  fig.ext = "jpg"
)
```

[\@drsimonj](https://twitter.com/drsimonj) here to make pretty histograms with ggplot2!

In this post you'll learn how to create histograms like this:

```{r init-example, echo = F, message = FALSE, warning = F}
library(ggplot2)

set.seed(070510)
d <- data.frame(x = rnorm(2000))

ggplot(d, aes(x, fill = cut(x, 100))) +
  geom_histogram(show.legend = FALSE) +
  scale_fill_discrete(h = c(180, 360), c = 120, l = 70) +
  theme_minimal() +
  labs(x = "Variable X", y = "n") +
  ggtitle("Histogram of X")
```

## The data

Let's simulate data for a continuous variable `x` in a data frame `d`:

```{r}
set.seed(070510)
d <- data.frame(x = rnorm(2000))

head(d)
```

## Basic Histogram

Create the basic ggplot2 histogram via:

```{r basic, message = F}
library(ggplot2)

ggplot(d, aes(x)) +
    geom_histogram()
```

## Adding Colour

Time to jazz it up with colour! The method I'll present was motivated by my answer to [this StackOverflow question](http://stackoverflow.com/questions/43795211/coloring-a-geom-histogram-by-gradient/43795338#43795338).

We can add colour by exploiting the way that ggplot2 stacks colour for different groups. Specifically, we fill the bars with the same variable (`x`) but `cut` into multiple categories:

```{r color1, message = F}
ggplot(d, aes(x, fill = cut(x, 100))) +
    geom_histogram()
```

What the...

Oh, ggplot2 has added a legend for each of the 100 groups created by cut! Get rid of this with `show.legend = FALSE`:

```{r color2, message = F}
ggplot(d, aes(x, fill = cut(x, 100))) +
    geom_histogram(show.legend = FALSE)
```

Not a bad starting point, but say we want to tweak the colours.

For a continuous colour gradient, a simple solution is to include `scale_fill_discrete` and play with the range of hues. To get your colours right, get familiar with the [hue scale](https://en.wikipedia.org/wiki/Hue).

For example, here we'll tweak the colours to range from blue to red:

```{r color3, message = F}
ggplot(d, aes(x, fill = cut(x, 100))) +
  geom_histogram(show.legend = FALSE) +
  scale_fill_discrete(h = c(240, 10))
```

Seems a little dark. Tweak chroma and luminance with `c` and `l`:

```{r color4, message = F}
ggplot(d, aes(x, fill = cut(x, 100))) +
  geom_histogram(show.legend = FALSE) +
  scale_fill_discrete(h = c(240, 10), c = 120, l = 70)
```

## Final touches

The final touches are to set the theme, add labels, and a title:

```{r touches, message = F}
ggplot(d, aes(x, fill = cut(x, 100))) +
  geom_histogram(show.legend = FALSE) +
  scale_fill_discrete(h = c(240, 10), c = 120, l = 70) +
  theme_minimal() +
  labs(x = "Variable X", y = "n") +
  ggtitle("Histogram of X")
```

Now have fun tweaking the colours!

```{r tweak, message = F}
p <- ggplot(d, aes(x, fill = cut(x, 100))) +
  geom_histogram(show.legend = FALSE) +
  theme_minimal() +
  labs(x = "Variable X", y = "n") +
  ggtitle("Histogram of X")

p + scale_fill_discrete(h = c(180, 360), c = 150, l = 80)

p + scale_fill_discrete(h = c(90, 210), c = 30, l = 50)
```

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).