---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.align = "center"
)
```

[\@drsimonj](https://twitter.com/drsimonj) here to explain how I used [ubeR](https://github.com/DataWookie/ubeR), an R package for the Uber API, to create this map of my trips over the last couple of years:

```{r init-example, message = FALSE, echo = F, warning = F}
library(ubeR)
library(tidyverse)

# This can't be run within RMarkdown.
#
# UBER_CLIENTID <- "XyXDcKSLKoVauGjc2fVfIZaKcitAOAzi"
# UBER_CLIENTSECRET <- "mgh6PZwyeFnuOMtQV7Pmu2tYk8QogApj30yFf05K"
# uber_oauth(UBER_CLIENTID, UBER_CLIENTSECRET)
#
#rides <- map_df(0:3, ~ uber_history(50, 50 * .))
#
# Instead, I ran it in a different project, saved the rides object, and put it
# in Rmd/ directory here to load
trips <- readRDS("trips.rds")

city_trips <- trips %>% 
  group_by(display_name) %>% 
  summarise(
    n = n(),
    long = mean(longitude),
    lat = mean(latitude)
  )

travel_paths <- trips %>% 
  # Find the from-to travel paths
  select(display_name) %>% 
  rename(from = display_name) %>% 
  mutate(to = lag(from)) %>% 
  filter(from != to) %>% 
  filter(!duplicated(.)) %>% 
  # Add coords for from city
  left_join(select(city_trips, -n), by = c("from" = "display_name")) %>% 
  rename(x = long, y = lat) %>% 
  # Add coords for to city
  left_join(select(city_trips, -n), by = c("to" = "display_name")) %>% 
  rename(xend = long, yend = lat)

long_min <- 120
long_max <- 300

world_map <- map_data(map = "world")

local_map <- world_map %>% 
  mutate(long = long + 360,
         group = group + max(group) + 1) %>% 
  rbind(world_map)

local_trips <- city_trips %>% 
  mutate(long = long + 360) %>% 
  rbind(city_trips)

local_paths <- travel_paths %>% 
  mutate(x = x + 360,
         xend = xend + 360) %>% 
  rbind(travel_paths) %>% 
  group_by(from, to) %>% 
  summarise(
    x = x[between(x, long_min, long_max)],
    xend = xend[between(xend, long_min, long_max)],
    y = y[1],
    yend = yend[1]
  )

ggplot(local_trips, aes(long, lat)) +
  geom_map(data = local_map, map = local_map,
           aes(map_id = region, x = long, y = lat),
           fill = "gray93", color = "gray85") +
  geom_curve(data = local_paths, color = "dodgerblue1",
             aes(x = x, y = y, xend = xend, yend = yend),
             arrow = arrow(angle = 20, type = "closed", length = unit(.18, "inches"))) +
  ggrepel::geom_text_repel(aes(label = display_name), force = 10, color = "black") +
  geom_point(aes(size = n), color = "dodgerblue3") +
  guides(size = "none", color = "none", alpha = "none") +
  coord_equal() +
  scale_x_continuous(limits = c(long_min, long_max)) +
  scale_y_continuous(limits = c(-50, 60)) +
  theme_void()
```

## Getting ubeR

The ubeR package, which I first heard about [here](https://www.r-bloggers.com/uber-a-package-for-the-uber-api/), is currently available on GitHub. In R, install and load it as follows:

```{r, eval = F}
# install.packages("devtools")  # Run to install the devtools package if needed
devtools::install_github("DataWookie/ubeR")  # Install ubeR
library(ubeR)
```

For this post I also use many of the tidyverse packages, so install and load this too to follow along:

```{r}
library(tidyverse)
```

## Setting up an app

To use ubeR and the uber API, you'll need an uber account and to register a new app. In a web browser, log into your uber account and head to [this page](https://developer.uber.com/dashboard/create). Fill in the details. Here's an example:

<< IMG OF REGISTER >>

Once created, under the *Authorization* tab, set the *Redirect URL* to http://localhost:1410/

<< IMG OF REDIRECT >>

Further down, under *General Scopes*, enable "profile" and "history":

<< IMG OF SCOPES >>

Back under the *settings* tab, under *Authentication*, take note of the "Client ID" and "Client Secret":

<< IMG OF AUTHENTICATION >>

In your R session, save these values as two variables:

```{r, eval = F}
UBER_CLIENTID <- "<< Your App's Client ID >>"
UBER_CLIENTSECRET <- "<< Your App's Client Secret >>"
```

## Using ubeR

Back in R, run the following:

```{r, eval = F}
uber_oauth(UBER_CLIENTID, UBER_CLIENTSECRET)
```

A web page will open asking you to log in to your uber account and permit the app to access your data. It will then say:

> Authentication complete. Please close this page and return to R.

We can now access our uber data via the R functions that access the uber API. For example `uber_me()` returns a vector of information contained in your profile.

```{r, eval = F}
me <- uber_me()
```
```{r, echo = F}
# Run above in R project and save variable as RDS so we can use it in this
# RMarkdown
me <- readRDS("me.rds")
```
```{r}
names(me)

cat("My uber profile name is", me$first_name, me$last_name)
```

## Accessing trip history

The function `uber_history` will return a data frame of portions of your trip history. As far as I can tell, you can extract up to 50 trips at a time. You can use `uber_history` to return one chunk of up to 50 trips. Alternatively, the following will read all of your trips into a data frame:

```{r, eval = F}
trips <- data.frame()
more_trips <- TRUE
off_set <- 0
while(more_trips) {
  new_trips <- uber_history(50, 50 * off_set)
  if (is.null(new_trips)) {
    more_trips <- FALSE
  } else {
    trips <- rbind(trips, new_trips)
    off_set <- off_set + 1
  }
}
```
```{r, echo = F}
# Read from saved rds
trips <- readRDS("trips.rds")
```
```{r}
trips %>% 
  mutate(longitude = "xx", latitude = "yy") %>%  # Just to hide values
  head

# How many trips have I taken?
nrow(trips)

# Time and place of my first trip
trips %>% 
  filter(start_time == min(start_time)) %>% 
  select(start_time, display_name)
```

## Getting the data into shape

For the map shown at the beginning, I created two data frames from my trip history:

- A data frame of how many trips I'd taken in each city, as well as an estimate of the city longitude and latitude. 
- A data frame of transitions from one city to another, along with their longitudes and latitudes.

### Trips per city

Calculating the trips per city is pretty straight forward. Given that we're using a world map, we can average the longitude and latitude for each city to get a good-enough position.

```{r}
city_trips <- trips %>% 
  group_by(display_name) %>% 
  summarise(
    n = n(),
    long = mean(longitude),
    lat = mean(latitude)
  )
city_trips
```

### City-to-city travel paths

The data comes with dates and is ordered from most recent to oldest. We can use this to approximately work out which cities I've gone from and to.

To get my travel paths, we find all occasions where the city for one trip doesn't match the city for the next trip. We then use the longitude and latitude from `city_trips` to create `x`, `y`, `xend`, and `yend` for the paths that we'll draw on the map.

```{r}
travel_paths <- trips %>% 
  # Find the from-to travel paths
  select(display_name) %>% 
  rename(from = display_name) %>% 
  mutate(to = lag(from)) %>% 
  filter(from != to) %>% 
  filter(!duplicated(.)) %>% 
  # Add coords for from city
  left_join(select(city_trips, -n), by = c("from" = "display_name")) %>% 
  rename(x = long, y = lat) %>% 
  # Add coords for to city
  left_join(select(city_trips, -n), by = c("to" = "display_name")) %>% 
  rename(xend = long, yend = lat)

travel_paths
```

## Plotting on the map (first attempt)

The following overlays the data onto the default map with the size of the points for each city determined by the number of trips I've taken there:

```{r, warning = F}
library(ggrepel)

world_map <-  map_data(map = "world")

ggplot(city_trips, aes(long, lat)) +
  # Plot map
  geom_map(data = world_map, map = world_map,
           aes(map_id = region, x = long, y = lat),
           fill = "gray93", color = "gray85") +
  # Plot travel paths
  geom_curve(data = travel_paths, color = "dodgerblue1",
             aes(x = x, y = y, xend = xend, yend = yend),
             arrow = arrow(angle = 20, type = "closed", length = unit(.18, "inches"))) +
  # Plot city names
  geom_text_repel(aes(label = display_name), force = 20, color = "black", size = 2.5) +
  # Plot city points 
  geom_point(aes(size = n), color = "dodgerblue3") +
  # Adjustments
  guides(size = "none", color = "none", alpha = "none") +
  coord_equal() +
  theme_void()
```

This is pretty good! However, I live in Australia, and I've only used uber internationally in the US.

## Plotting on a localized map

To display my trip information a little more clearly, the following creates a local version of the map:

```{r, warning = F}
long_min <- 120
long_max <- 300

local_map <- world_map %>% 
  mutate(long = long + 360,
         group = group + max(group) + 1) %>% 
  rbind(world_map)

local_trips <- city_trips %>% 
  mutate(long = long + 360) %>% 
  rbind(city_trips)

local_paths <- travel_paths %>% 
  mutate(x = x + 360,
         xend = xend + 360) %>% 
  rbind(travel_paths) %>% 
  group_by(from, to) %>% 
  summarise(
    x = x[between(x, long_min, long_max)],
    xend = xend[between(xend, long_min, long_max)],
    y = y[1],
    yend = yend[1]
  )

ggplot(local_trips, aes(long, lat)) +
  geom_map(data = local_map, map = local_map,
           aes(map_id = region, x = long, y = lat),
           fill = "gray93", color = "gray85") +
  geom_curve(data = local_paths, color = "dodgerblue1",
             aes(x = x, y = y, xend = xend, yend = yend),
             arrow = arrow(angle = 20, type = "closed", length = unit(.18, "inches"))) +
  ggrepel::geom_text_repel(aes(label = display_name), force = 10, color = "black") +
  geom_point(aes(size = n), color = "dodgerblue3") +
  guides(size = "none", color = "none", alpha = "none") +
  coord_equal() +
  scale_x_continuous(limits = c(long_min, long_max)) +
  scale_y_continuous(limits = c(-50, 60)) +
  theme_void()
```

A brief explanation of my trips:

- I live in Sydney, Australia, so most of my uber trips are there.
- I frequently travel to Melbourne to visit family.
- I often work in Adelaide, but usually hire a car, so don't use uber much while I'm there.
- I did a summer internship at ETS in Princeton, so that included my trip through New Jersey and Philadelphia.
- After finishing my Ph.D. I took a trip around the US, which involved Dallas, Miami, New Orleans, and -- my inner nerd couldn't resist -- a conference in Chicago before coming home.

It's not a complete travel history, but pretty cool to see just through my use of uber!

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).