---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 5,
  fig.width = 6,
  fig.align = "center",
  fig.ext = "tiff"
)
```

[\@drsimonj](https://twitter.com/drsimonj) here to make pretty scatter plots of correlated variables with ggplot2!

We'll learn how to create plots that look like this:

```{r init-example, message = FALSE, warning = F, echo = F}
library(ggplot2)

set.seed(170513)
n <- 2000
d <- data.frame(a = rnorm(n))
d$b <- .5*d$a + rnorm(n)

# Add first principal component
d$pc <- predict(prcomp(~a+b, d))[,1]

# Fit bivariate density
density_fit <- MASS::kde2d(d$a, d$b)

# Add estimated density for each point
d$density <- fields::interp.surface(
  density_fit, d[,c("a", "b")])

ggplot(d, aes(a, b, color = pc, alpha = 1/density)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e") +
  scale_alpha(range = c(.25, .6))
```

## Data

In a data.frame `d`, we'll simulate two correlated variables `a` and `b` of length `n`:

```{r}
set.seed(170513)
n <- 200
d <- data.frame(a = rnorm(n))
d$b <- .4 * (d$a + rnorm(n))

head(d)
```

## Basic scatter plot

Using ggplot2, the basic scatter plot (with `theme_minimal`) is created via:

```{r}
library(ggplot2)

ggplot(d, aes(a, b)) +
  geom_point() +
  theme_minimal()
```

## Shape and size

There are many ways to tweak the `shape` and `size` of the points. Here's the combination I settled on for this post:

```{r}
ggplot(d, aes(a, b)) +
  geom_point(shape = 16, size = 5) +
  theme_minimal()
```

## Color

We want to color the points in a way that helps to visualise the correlation between them.

One option is to `color` by one of the variables. For example, color by `a` (and hide legend):

```{r}
ggplot(d, aes(a, b, color = a)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal()
```

Although it's subtle in this plot, the problem is that the color is changing as the points go from left to right. Instead, we want the color to change in a direction that characterises the correlation - diagonally in this case.

To do this, we can color points by the **first** principal component. Add it to the data frame as a variable `pc` and use it to color like so:

```{r}
d$pc <- predict(prcomp(~a+b, d))[,1]

ggplot(d, aes(a, b, color = pc)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal()
```

Now we can add color, let's pick something nice with the help of the `scale_color_gradient` functions and some nice hex codes (check out [color-hex](http://www.color-hex.com/) for inspriation). For example:

```{r}
ggplot(d, aes(a, b, color = pc)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```

## Transparency

Now it's time to get rid of those offensive mushes by adjusting the transparency with `alpha`.

We could adjust it to be the same for every point:

```{r}
ggplot(d, aes(a, b, color = pc)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE, alpha = .4) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```

This is fine most of the time. However, what if you have many points? Let's try with 5,000 points:

```{r}
# Simulate data
set.seed(170513)
n <- 5000
d <- data.frame(a = rnorm(n))
d$b <- .4 * (d$a + rnorm(n))

# Compute first principal component
d$pc <- predict(prcomp(~a+b, d))[,1]

# Plot
ggplot(d, aes(a, b, color = pc)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE, alpha = .4) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```

We've got another big mush. What if we take `alpha` down really low to .05?

```{r}
ggplot(d, aes(a, b, color = pc)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE, alpha = .05) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```

Better, except it's now hard to see extreme points that are alone in space.

To solve this, we'll map `alpha` to the **inverse** point density. That is, turn down `alpha` wherever there are lots of points! The trick is to use bivariate density, which can be added as follows:

```{r}
# Add bivariate density for each point
d$density <- fields::interp.surface(
  MASS::kde2d(d$a, d$b), d[,c("a", "b")])
```

Now plot with `alpha` mapped to `1/density`:

```{r}
ggplot(d, aes(a, b, color = pc, alpha = 1/density)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```

You can see that distant points are now too vibrant. Our final fix is to use `scale_alpha` to tweak the alpha range. By default, this range is 0 to 1, making the most distant points have an alpha close to 1. Let's restrict it to something better:

```{r}
ggplot(d, aes(a, b, color = pc, alpha = 1/density)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal() +
  scale_color_gradient(low = "#0091ff", high = "#f0650e") +
  scale_alpha(range = c(.05, .25))
```

Much better! No more mushy patches or lost points.

## Bringing it together

Here's a complete example with new data and colors:

```{r}
# Simulate data
set.seed(170513)
n <- 2000
d <- data.frame(a = rnorm(n))
d$b <- -(d$a + rnorm(n, sd = 2))

# Add first principal component
d$pc <- predict(prcomp(~a+b, d))[,1]

# Add density for each point
d$density <- fields::interp.surface(
  MASS::kde2d(d$a, d$b), d[,c("a", "b")])

# Plot
ggplot(d, aes(a, b, color = pc, alpha = 1/density)) +
  geom_point(shape = 16, size = 5, show.legend = FALSE) +
  theme_minimal() +
  scale_color_gradient(low = "#32aeff", high = "#f2aeff") +
  scale_alpha(range = c(.25, .6))
```

## Sign off

Thanks for reading and I hope this was useful for you.

For updates of recent blog posts, follow [\@drsimonj](https://twitter.com/drsimonj) on Twitter, or email me at <drsimonjackson@gmail.com> to get in touch.

If you'd like the code that produced this blog, check out the [blogR GitHub repository](https://github.com/drsimonj/blogR).