---
title: "Proportions with mean()"
author: "Simon A Jackson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "post-"
)
```

One of the most common tasks I want to do is calculate the proportion of observations (e.g., rows in a data set) that meet a particular condition. For example, what is the proportion of missing data, or people over the age of 18?

There is a suprisingly easy solution to handle this problem: by combining boolean vectors and `mean()`.


## Step 1: creating a boolean vector

We start with boolean vectors, which is a vector that is `TRUE` whenever our observation meets our condition, or `FALSE` whenever it's not. We create this boolean vector by submitting our observations to some sort of conditional statement (or relevant function like `is.na()`). Let's take a look at a few examples:

```{r}
x <- letters[1:10]
x == "b"  # return a boolean vector which is TRUE whenver x is "b"

x <- 1:10
x > 5  # TRUE whenever x is greater than 5

x > 5 & x %% 2 == 0  # TRUE when x > 5 AND divisible by 2

x <- c(1, 2, NA, 4)
is.na(x) # TRUE when x is a missing value
```

If you're unsure with how the above works, take a look at this page on [R Programming Operators](http://www.programiz.com/r-programming/operator).

With this under our belt, it seems simple enough to create a boolean vector that tells us when our observations meet some condition (`TRUE`) or not (`FALSE`).

## Step 2: calculating the proportion of TRUE

From this point, all we need to do is wrap our conditional statment inside `mean()`:

```{r}
x <- 1:10
mean(x > 5)  # proportion of values in x greater than 5
```

How/why does this work? If you take a look at the help page with `?mean()`, you'll read that the arguement `x` can be a logical vector. But what does this mean. Well, when you use a boolean vector, `mean()` first converts it to a numeric vector. This means that every `TRUE` becomes `1`, and every `FALSE` becomes `0`:

```{r}
x <- 1:10
as.numeric(x > 5)
```

It then computes the mean of these 1's and 0's. At this point, you just need to think a little. How is the mean calculated? Well, it's the sum of all the values, divided by their length? So the sum of a vector of 1's and 0's will be the total number of 1's! Divided by the length then gives you the proportion. As a side note, you might realise that you can use `sum()` instead of `mean()` if you want to calculate the frequency. Let's break this right down:

```{r}
x <- 1:10
x

test <- x > 5
test

as.numeric(test)

sum(test)

length(test)

sum(test) / length(test)

mean(test)
```

## Some useful examples

At this point, we can apply this to all sorts of problems. Here are some examples using the mtcars data set:

```{r}
d <- mtcars
head(d)

# Proportion of rows (cars) with cyl == 6 (6 cylinders)
mean(d$cyl == 6)

# Proportions of rows (cars) with hp > 250 (horsepower over 200)
mean(d$hp > 250)

# Proportion of cars with 8-cylinders and that get more than 15 Miles/(US) gallon
mean(d$cyl == 8 & d$hp > 15)
```


